---
title: "WHO Suicide Statistics Final"
author: "Jaquelin Martinez"
date: "December 17, 2018"
output: word_document
---

##Data
The dataset used for this analysis is merged data from WHO suicide statistics (https://www.kaggle.com/szamil/who-suicide-statistics) and World Bank data that was accessed through R by a package (“WDI”). The WHO suicide statistics had data on raw number of suicides and population, divided into age groups and gender, for certain countries for 1987-2016. The WDI provided data on GDP per capita, PPP (constant 2011 international $), income level, and region for countries in 1992-2016. The final merged dataset contains only countries and year common to both dataset.

My main question of exploration for merging the datasets is to see if there is a relationship between GDP per capita (PPP) and suicide rates (# suicides per 100,000 people in the population). I created the suicide rate to make the numbers comparable across age groups and countries with less population.

I would also like to explore differeces in suicide rates between man and women and through the years in our data set. 


```{r, cache=TRUE}
source("final_processing.R")
source("final_exploring.R")
source("final_aggregate.R")
source("final_regression.R")
```



```{r, cache=TRUE} 
df <- merged_who_wdi()
```


What are the countries with the highest suicide rate (or the highest raw number of suicides) for each year in our dataset?
```{r, cache=TRUE} 
source("final_aggregate.R")

top_rate_year(df)
top_raw_year(df)

```


Given these results, some countries of interest are: Findland, Hungary, Russian Federation, Lithuania, United States, Thailand. 


```{r, cache=TRUE} 
graph_by_country(df, "Lithuania", 1996)
graph_by_country(df, "Russian Federation", 1994)
graph_by_country(df, "United States of America", 2015)
graph_by_country(df, "Thailand", 2016)
```
These graphs show the great disparities in suicide rates between men and women. 

```{r, cache=TRUE} 
time_trend(df, "Lithuania")
time_trend(df, "Hungary")
time_trend(df, "Finland")
time_trend(df, "United States of America") 
time_trend(df, "Russian Federation") 
time_trend(df, "Thailand") 
```


To check if there is a visible relationship, we can plot country-level suicide rates and GDP(PPP) data. 
```{r, cache=TRUE} 
plot_aggregate(df) 
```

There is no visible relationship, given this data, between GDP(PPP- constant 2011 int $) and suicide rates.  The plots are divided by decades. The 90s reflect a slightly positive relationship between GDP and suicide rates.

We can verify that with a simple regression on GDP (PPP) on the suicide rate. 

Our second regression model, a multivarate_regression, is much better model for creating a predicted suicide rate, as it adjusts for country, age group, gdp, and gender. Though the coefficients are not shown, we can see that the adjusted R-squared is: 0.5213. 
. 

```{r, cache=TRUE} 
simple_regression(df)
multivarate_regression(df)
```
