---
title: "Final Project"
author: "Ziyun Wang"
date: "12/15/2018"
output: pdf_document
---

# Introduction and Background
The sinking of the RMS Titanic is one of the most infamous shipwrecks in history.  On April 15, 1912, during her maiden voyage, the Titanic sank after colliding with an iceberg, killing 1502 out of 2224 passengers and crew. This sensational tragedy shocked the international community and led to better safety regulations for ships.

One of the reasons that the shipwreck led to such loss of life was that there were not enough lifeboats for the passengers and crew. Although there was some element of luck involved in surviving the sinking, some groups of people were more likely to survive than others, such as women, children, and the upper-class.

# Data Sources & Variables
##Data Source:
**Kaggle** (https://www.kaggle.com/c/titanic/data)


##Response Variable: 

**Survival** (0 = died, 1 = survived)


##Expanatory Variables:

**Name**: Name of Passenger

**Pclass**: Ticket Class of Travel (1 = First, 2 = Second, 3 = Third)

**Age**: Passenger Age in years (continuous)

**Sex**: Gender (Male & Female)

**Sibsp**: Number of Sibing/Spouse on aboard (continuous)

**Parch**: Number of Parent/Child on aboard (continuous)

**Ticket**: Ticket Number

**Fare**: Price of the Ticket (continuous)

**Cabin**: Cabin Number where Passenger Stayed

**Embarked**: Port in which Passenger Embarked (C - Cherbourg, S - Southampton, Q = Queenstown)

# Question of Interest
**How do factors such as Class, Gender, Age, Title, Number of Family Members on aboard affect one's chance of survival on Titanic?**


# Data Analysis

**Data and R setup**
```{r, message=F, warning=F}
library(ggplot2, quietly = TRUE)
library(dplyr, quietly = TRUE)
install.packages('ggthemes',repos = "http://cran.us.r-project.org")
library(ggthemes, quietly = TRUE)
install.packages('corrplot',repos = "http://cran.us.r-project.org")
library(corrplot, quietly = TRUE)
titanic <- read.csv("titanic.csv", header = T, stringsAsFactors = F)
```

##**Summary of Dataset**
```{r}
head(titanic)[1:5,]
summary(titanic)
```

##**Number of Survivals with One Factor Considered**
```{r}
survival_factor <- function(df, variable)
{
  survive <- df %>%
    mutate(Class = ifelse(Survived == 0, 'Died', 'Survived')) %>%
     group_by_(variable, 'Class') %>%
    summarise(count = n())
  
  p <- ggplot(survive, aes(y=survive$count, x=survive[[variable]], 
                           color=survive$Class, 
                           fill=survive$Class)) 
  p <- p + geom_bar(stat="identity", position="dodge") + labs(color="Class", fill="Class", y="Number of Passenger",x=variable)
  print(p)
}

```

\newpage

**Number of Survivals with Class that Passengers Stayed in**
```{r}
survival_factor(titanic, 'Pclass')
```


**Insight**
As we can see from the graph, passengers who stayed in the third class had a much lower chance to survive compared with the first and secons class passengers.

\newpage

**Number of Survivals with Gender of Passengers**
```{r}
survival_factor(titanic, 'Sex')
```


**Insight**
As we can see from the graph, female passengers had a higher chance to survive than male passengers on Titanic. This might result from the cultural behavior, which is allowing women children to leave first when something bad happens.

##**Average Ticket Price of Each Class**
```{r}
class_fare <- function(df) {
  m <- rep(NA, 3)
  for (i in 1:length(m)) {
    class_df <- dplyr::filter(df, Pclass == i)
    m[i] <- mean(class_df$Fare)
  }
  fare_df <- data.frame(Class=1:3, Avg.Fare=m, stringsAsFactors = F)
  
  return (fare_df)
}

class_fare(titanic)
```


##**Number of Survivals with Two Factor Considered**
```{r}
survival_two_factors <- function(df, f1, f2) {
  ggplot(df, aes(df[[f1]], fill = factor(df$Survived))) + 
    geom_bar(stat = "count")+
    theme_few() +
    xlab("Pclass") +
    facet_grid(.~df[[f2]])+
    ylab("Count") +
    scale_fill_discrete(name = "Survived") + 
    ggtitle("Pclass vs Sex vs Survived")
}
```


**Number of Survivals with Gender of Passengers and Class Stayed In**
```{r}
survival_two_factors(titanic, 'Pclass', 'Sex')
```


**Insight**

From the graph with two factors considered, we can say that most females from the first and second class suvived. Sadly, most males in the thrid class died in the shipwreck of Titanic.

\newpage

##**Number of Survivals with Three Factor Considered**
```{r}
survival_three_factors <- function(df, Age, Sex, Pclass){
  p <- ggplot(df, aes(x = Age, y = Sex)) + 
    geom_jitter(aes(colour = factor(Survived))) + 
    theme_few()
  p <- p + theme(legend.title = element_blank())+
    facet_wrap(~Pclass) 
  p <- p + labs(x = "Age", y = "Sex", title = "Survivor factors: Class vs Sex vs Age")
  p <- p + scale_fill_discrete(name = "Survived") + 
    scale_x_continuous(name="Age",limits=c(0, 81))
  print(p)
}
```


**Number of Survivals with Gender and Age of Passenger, and Class Stayed In**
```{r}
survival_three_factors(titanic, 'Age', 'Sex', 'Pclass')
```


**Insight**

In this graph, which age factor was incorporated, it gives us an idea how age was affecting the survival chance.

\newpage

##**Number of Survivals with the Title of the Passenger**
```{r}
survival_title <- function(df) {
  df$Title <- gsub('(.*, )|(\\..*)', '', df$Name)
  
  officer <- c('Capt', 'Col', 'Don', 'Dr', 'Major', 'Rev')
  royalty <- c('Dona', 'Lady', 'the Countess','Sir', 'Jonkheer')
  
  # Reassign titles
  df$Title[df$Title == 'Mlle']        <- 'Miss' 
  df$Title[df$Title == 'Ms']          <- 'Miss'
  df$Title[df$Title == 'Mme']         <- 'Mrs' 
  df$Title[df$Title %in% royalty]  <- 'Royalty'
  df$Title[df$Title %in% officer]  <- 'Officer'
  
  df$Surname <- sapply(df$Name,  
                            function(x) strsplit(x, split = '[,.]')[[1]][1])
  p <- ggplot(df, aes(Title,fill = factor(Survived))) +
    geom_bar(stat = "count")+ xlab('Title') + ylab("Count") 
  p <- p + scale_fill_discrete(name = " Survived") + 
    ggtitle("Title vs Survived")+
    theme_few()
  print(p)
}
survival_title(titanic)
```

\newpage
 
**Insight**

In this graph, we have a rough idea on how one's title affects one's chance to survive from the shipwreck of Titanic. All passengers who were royal memebers survived. Also, married women had a high chance to survive, which might because their children were with them and so they were identified as the first priority to leave the ship.


##**Number of Survivals with the Family Size of the Passenger**
```{r}
survival_familysize <- function(df) {
  df$Fsize <- df$SibSp + df$Parch + 1
  
  p <- ggplot(df, aes(x = Fsize, fill = factor(Survived))) +
    geom_bar(stat='count', position='dodge') 
  p <- p + scale_x_continuous(breaks=c(1:11)) + xlab('Family Size') +
    ylab("Count")
  p <- p + theme_few()+scale_fill_discrete(name = "Survived") 
  p <- p + ggtitle("Family Size vs Survived")
  print(p)
}
survival_familysize(titanic)
```


**Insight**

From this graph, it shows us how number of family members one has would affect one's chance of survival from the shipwreck. People with more than 4 family members on aboard with them were less likely to survive, which might becuase they used more time looking for their family while the tragedy happened. However, people has no family members also had a lower chance to survive. Surpirsingly, people with one to two family members had the highest chance to survive among others.


##**Correlation Maytix of Number of Survivals and All factors**
```{r}
#detach("package:dplyr", unload=TRUE)
library(plyr)
survival_correlation <- function(df) {
  df <- titanic
  df$Fsize <- df$SibSp + df$Parch + 1
  df$FsizeD[df$Fsize == 1] <- 'Alone'
  df$FsizeD[df$Fsize < 5 & df$Fsize > 1] <- 'Small'
  df$FsizeD[df$Fsize > 4] <- 'Big'
  df$Title <- gsub('(.*, )|(\\..*)', '', df$Name)
  
  officer <- c('Capt', 'Col', 'Don', 'Dr', 'Major', 'Rev')
  royalty <- c('Dona', 'Lady', 'the Countess','Sir', 'Jonkheer')
  
  # Reassign titles
  df$Title[df$Title == 'Mlle']        <- 'Miss' 
  df$Title[df$Title == 'Ms']          <- 'Miss'
  df$Title[df$Title == 'Mme']         <- 'Mrs' 
  df$Title[df$Title %in% royalty]  <- 'Royalty'
  df$Title[df$Title %in% officer]  <- 'Officer'
  
  corr_data <- df
  corr_data$Sex <- revalue(corr_data$Sex, 
                           c("male" = 1, "female" = 2))
  corr_data$Title <- revalue(corr_data$Title, 
                             c("Mr" = 1, "Master" = 2,"Officer" = 3, 
                               "Mrs" = 4,"Royalty" = 5,"Miss" = 6))
  corr_data$FsizeD <- revalue(corr_data$FsizeD, 
                              c("Small" = 1, "Alone" = 2, "Big" = 3))
  corr_data$FsizeD <- as.numeric(corr_data$FsizeD)
  corr_data$Sex <- as.numeric(corr_data$Sex)
  corr_data$Title <- as.numeric(corr_data$Title)
  corr_data$Pclass <- as.numeric(corr_data$Pclass)
  corr_data$Survived <- as.numeric(corr_data$Survived)
  
  corr_data <-corr_data[,c("Survived", "Pclass", "Sex", 
                           "FsizeD", "Fare", "Title")]
  
  str(corr_data)
  mcorr_data <- cor(corr_data)
  
  corrplot(mcorr_data,method="circle")
}

survival_correlation(titanic)
```

# Conclusion
To sum up, the passenger's gender, title and the class stayed in play influencial roles in determining tht passenger's chance to survive from the shipwreck of the Titanic. 